---
title: "intro_ggplot2"
output: html_document
---

### Set up

First things first, change these settings

Tools -> Global options -> R Markdown -> Uncheck "Show output inline for all R Markdown documents"


### R Packages
- "Packages" are add-ons to plain R

- These installation commands are commented out because I've already installed these packages in this workspace

- If you're on your own computer, delete the # in front of each one and run them all

- To run a command in Rstudio, have your cursor on the line you want to run and either hit the Run green arrow or hit ctrl-enter


- To run a whole gray chunk (like the one below), click the green arrow at the top of the chunk

- Things in chunks are code, things outside of chunks are for text and notes

```{r packages}

#Necessary packages
#install.packages("tidyverse") # Tools for working with data
#install.packages("dslabs") # Our example dataset

#Less critical packages for single examples, don't worry about if they won't install
#install.packages("ggridges") # A type of plot
#install.packages("cowplot") # Tools for changing plot aesthetics
#install.packages("skimr") # For summary statistics of a data table
#install.packages("gganimate") # For animations
#install.packages("gifski") # For animations
#install.packages("png") # For animations
#install.packages("plotly") # For interactivity example


library(tidyverse)
library(dslabs)
library(ggridges)
library(cowplot)
library(skimr)
library(gganimate)
library(plotly)
library(RColorBrewer)
theme_set(theme_cowplot())
```

- We'll start with a basic plot using a built-in example dataset (called mpg)

- Run this the code in the chunk below to get a preview of this data in the Console

```{r}
mpg
```


- To see the whole dataset, use the function View() to pop it into a preview tab
```{r}
View(mpg)
```

- To get a description of your data
```{r}
skim(mpg)
```


- The template for making a graph is:

ggplot(data = DATASETNAME, mapping = aes(x = XCOLUMN, y = YCOLUMN)) + 
   geom_TYPEOFGRAPH() 
   
Where things in all caps are variables that you change. aes stands for "aesthetics", this is where you tell ggplot what column in your data is the x axis, what is the y axis
   
For the mpg dataset, we can plot a scatterplot of the columns for engine displacement `displ` vs highway miles per gallon `hwy`

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point()
```


You can also define what column describes color

We can color our scatterplot by what class of car each observation is
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = class)) + 
  geom_point()

```

Or point shape
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = class, shape = class)) + 
  geom_point()
```

We can add another geom to fit a smooth curve to our data
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() +
  geom_smooth()
```

We can break into one subgraph per vehicle class
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() +
  facet_wrap(~class)
```

Or animate based on a column
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = cyl)) + 
  geom_point() +
  transition_manual(class) +
  labs(title = 'Car class: {current_frame}')

```

Or create an interactive plot
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, label = class)) + 
  geom_point() 

ggplotly()
```

Pretty much any characteristic of a plot you can change by adding on more lines to your ggplot statement
Googling a template like "ggplot change color scale", or "ggplot change x axis range" will generally lead to answers


# For the code-along portion of this, I'll be replicating a famous plot of measles incidence before and after the vaccine was introduced in 1963

First, the dslabs package contains the dataset used for this infographic
```{r}
data(us_contagious_diseases)
us_contagious_diseases <- as_tibble(us_contagious_diseases)
```

Take a look
```{r}
us_contagious_diseases
```


Skim shows that there are 7 diseases and data on 51 states (including D.C.)
```{r}
skim(us_contagious_diseases)
```

Introducing the pipe (%>%) aka "and then" for exploratory data analysis
With the pipe, you can string together analysis steps
```{r}
us_contagious_diseases %>% 
  select(disease) %>%
  unique

# Take the data set, and then...
# select a specific column, and then ...
# Get unique values of that column

```


Get just rows for Measles using a filter, and save as a new variable called measles_dat
```{r}
measles_dat <- us_contagious_diseases %>%
  filter(disease == "Measles")

```

Add on a new column for rate by using the mutate function
```{r}
measles_dat <- measles_dat %>%
  mutate(rate = (count / population) * 10000 * (52 / weeks_reporting))
 
measles_dat
```

Now set up the ggplot
```{r}
ggplot(data = measles_dat, aes(x = year, y = state, fill = rate)) + 
  geom_tile()
```

That's one ugly plot, now to make it nice
Start with changing the color scale
```{r}
ggplot(data = measles_dat, aes(x = year, y = state, fill = rate)) + 
  geom_tile() +
  scale_fill_gradientn(colors = c("white", "blue"))
```

Missing data shows as grey blocks, let's make them less obtrusive with na.value

How was missing data handled on the original plot?
```{r}
ggplot(data = measles_dat, aes(x = year, y = state, fill = rate)) + 
  geom_tile() +
  scale_fill_gradientn(colors = c("white", "blue"), na.value = "grey95")
```

We need to add a vertical line where vaccine was introduced
```{r}
ggplot(data = measles_dat, aes(x = year, y = state, fill = rate)) + 
  geom_tile() +
  scale_fill_gradientn(colors = c("white", "blue"), na.value = "grey95") +
  geom_vline(xintercept = 1963, col = "black") 

```

Now add accessory text
```{r}

ggplot(data = measles_dat, aes(x = year, y = state, fill = rate)) + 
  geom_tile() +
  scale_fill_gradientn(colors = c("white", "blue"), na.value = "grey95", name = "Cases per\n10,000 people") +
  geom_vline(xintercept = 1963, col = "black") +
  ggtitle("Measles") +
  ylab("") +
  xlab("") +  
  annotate(geom = "text", x = 1963 + 1, y = 50.5, label = "Vaccine introduced", size = 5, hjust = 0) 
```


In original, states were ordered A-Z top to bottom
We can use fct_ tools to change plot ordered
```{r}
ggplot(data = measles_dat, aes(x = year, y = fct_rev(state), fill = rate)) + 
  geom_tile() +
  scale_fill_gradientn(colors = c("white", "blue"), na.value = "grey95", name = "Cases per\n10,000 people") +
  geom_vline(xintercept = 1963, col = "black") +
  ggtitle("Measles") +
  ylab("") +
  xlab("") +  
  annotate(geom = "text", x = 1963 + 1, y = 50.5, label = "Vaccine introduced", size = 5, hjust = 0) 
```


```{r}




```

